---
title: "Chicago Marathon Analysis and the Impact of Weather on Performance"
author: "Miles Maloney"
institute: "STAT385"
date: "August 7, 2026"
format: 
  revealjs: 
    theme: default
    transition: slide
    footer: "Chicago Marathon Weather Impact Analysis"
editor: visual
#note: when hitting Render it does update on http://localhost:7093/#/title-slide
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

library(shiny)
library(tidyverse)
library(plotly)
library(DT)
library(broom)
library(leaflet)
library(sf)
#data
gpx_map_2000 <- st_read("2000 Chicago Marathon Route.gpx", layer = "tracks", quiet = TRUE)
gpx_map_2025 <- st_read("2025 Chicago Marathon Route.gpx", layer = "tracks", quiet = TRUE)
runner_data <- read_csv("STAT385 Project (2000 to 2025 results).csv")
weather_data <- read_csv("STAT385 Project (weather working set).csv")

#data wrangling
set.seed(1)
runners_finished <- runner_data %>%
  drop_na(finish_time_seconds) %>%
  filter(finish_time_seconds <= 23400) %>%
  group_by(year) %>%
  mutate(quartile=ntile(finish_time_seconds, 4)) %>%
  group_by(year, quartile) %>%
  slice_sample(n = 1000) %>%
  ungroup()

weather_needed <- weather_data %>%
  mutate(year = year(time), hour = hour(time))

racing_weather <- weather_needed %>%
  filter(hour >= 7 & hour <= 15) %>%
  group_by(year) %>%
  summarize(
    avg_race_temp_f = mean(temperature_2m_f, na.rm = TRUE), 
    avg_humidity_pct = mean(relative_humidity_2m_pct, na.rm = TRUE),
    total_precip_in = sum(precipitation_inch, na.rm = TRUE),
    .groups = "drop"
  )
final_data <- runners_finished %>%
  left_join(racing_weather, by = "year")
top_1percent <- runner_data %>%
  drop_na(finish_time_seconds) %>%
  filter(finish_time_seconds <= 23400) %>%
  group_by(year) %>%
  mutate(percentile = ntile(finish_time_seconds, 100)) %>%
  filter(percentile == 1) %>% 
  ungroup() %>%
  left_join(racing_weather, by ="year")

spread_data <- bind_rows(
  final_data %>% mutate(runner_group = "General Sample"),
  top_1percent %>% mutate(runner_group = "Top 1% Elites")
)

general_model <- lm(finish_time_seconds ~ avg_race_temp_f, data = final_data)
elite_model <- lm(finish_time_seconds ~ avg_race_temp_f, data = top_1percent)

general_r2 <- summary(general_model)$adj.r.squared
elite_r2 <- summary(elite_model)$adj.r.squared

gen_ci <- tidy(general_model, conf.int = TRUE, level = 0.95) %>% filter(term == "avg_race_temp_f")
elite_ci <- tidy(elite_model, conf.int = TRUE, level = 0.95) %>% filter(term == "avg_race_temp_f")

gen_slope <- gen_ci$estimate
elite_slope <- elite_ci$estimate  

```

# Motivation

## Why This Topic Matters

::: incremental
- The Chicago Marathon is one of the largest races in the world.
- It is consistently held annually on the second Sunday of October.
- Over 50,000 runners participate in it in recent years.
- Runners spend months or years training for the Marathon.
- Weather on the race day is unpredictable.
- Understanding how runners are sensitive to the temperature conditions helps them set personal goals and helps marathon directors prepare for medical needs during the race.
:::

## Personal Connection

::: incremental
- I was born and raised in Chicago and have been an avid runner for many years.
- I like to investigate how external factors impact performance.
- I have never run the marathon but have volunteered for it many times, manning a water station.
- The route takes the runners right past my former high school, the entire cross country team would gather to cheer on our coach as he ran.
:::

## Research Questions

::: incremental
- What weather element impacts runners the most on marathon days?
- What groups of runners are impacted the most by weather on marathon days?
- How much of an impact does weather have on finish times?
- What are the ideal conditions for running the Chicago Marathon?
:::

# Data Description

## Data Source

- The Marathon results were collected from Kaggle as a CSV file. It contained results from all runners from 2000-2025, and had over 900,000 finishers.
- The hourly weather data was collected from Open-Meteo as CSV files, which had high-resolution data from national weather services.
- Every year, the weather data was selected from the race window between 7am and 3pm. 7am was the earliest possible starting time, and 3pm was the latest possible finishing time.

## Variables

- The dependent variable is finish time. It was collected in seconds, but I converted it to hours.
- The independent variables are from the weather data, which are: Average Temperature (°F), Apparent Temperature (°F), Relative Humidity (%), and Total Precipitation (inches).
- The categorical variables are gender, age group, finish time quartiles, general sample of all runners, and top 1% of finishers.

## Data Size

- The raw data from race results included over 930,000 finishers across 25 years.\
- I took a stratified random sample of 1,000 runners in each quartile from each year. This was to make sure that each sample had the same size and to ensure efficiency.
- The top 1% of runners' data from each year was collected and put in a different data set.

## Data Limitations

- Certain runner's heats started at different times.\
- There is no data on runner's training volume, diet, or potential personal challenges that could impact finish time.
- The model uses the average of the weather from 7am to 3pm.

# Methods

## Data Cleaning
- Filtered the data by the finishing time, so that all times were less than the Chicago Marathon time limit of 6 hours and 30 minutes (15-minute per mile pace).
- Removed runners who did not finish the race.
- Merged the hourly weather averages to the runner results based on year and start time of different groups.
```{r}
#| eval: false
# Example of filtering and data cleaning logic:
runners_finished <- runner_data %>%
  drop_na(finish_time_seconds) %>%
  filter(finish_time_seconds <= 23400)
```

## Exploratory Analysis

::: incremental
- The marathon route changes a little from year to year due to factors such as construction, traffic patterns, and other events happening at the same time. 
- I wanted to see how significant the changes might be, so I used geographic data and GPX files to illustrate the route from the first year of the dataset and compare it to the last year of the dataset.
:::

## Exploratory Analysis
- I plotted the overall distribution of the marathon finishing times.
- The distribution of marathon times is a broad bell shape. There are clusters before the finishing times of 3 hours, 4 hours, and 5 hours due to common goals that runners share.\

## Marathon Finish Time Distribution

```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 7.5
#| fig-width: 10
#| out-width: "100%"
ggplot(final_data, aes(x=finish_time_seconds/3600)) +  
  geom_histogram(aes(y=after_stat(density)), bins=100, color = "gray", fill="lightgray") +  
  geom_density(color ="blue", linewidth = 1) +
  labs(
    title = "Marathon Finish Times Distribution (2000-2025)",
    x = "Finishing Time (hours)",
    y = "Density"
  ) +  
  theme_minimal()
```

<div style="font-size: 75%;">
- This marathon finishing time distribution density plot is moderately right skewed.
- The majority of runners finish within a time range of 4 to 5 hours.
- There is 1 large bump present which indicates a common finishing time goal of under 4 hours.
</div>

## Interactive Course Map Comparison (2000 and 2025)


```{r}
#| echo: false
#| message: false
leaflet(height = "420px", width = "100%") %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolylines(data = gpx_map_2000, color = "orange", weight = 5, opacity = 0.9, 
               label = "2000 Course", group = "2000 Course") %>%
  addPolylines(data = gpx_map_2025, color = "blue", weight = 5, opacity = 0.9, 
               label = "2025 Course", group = "2025 Course") %>%
  addLayersControl(
    overlayGroups = c("2000 Course", "2025 Course"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

## Statistical Methods

- I used the linear regression model lm to find the relationship between weather factors and finishing times.
- I used the broom package to calculate the slopes and R\^2 values.
#| eval: false
# Regression model fitting code
general_model <- lm(finish_time_seconds ~ avg_race_temp_f, 
                    data = final_data)
                    
# Visualizations

## Thermal Sensitivity
This plot illustrates the contrast in heat sensitivity between the general runner sample and the top 1% elite runners. This shows how temperature changes impact finish times across different skill levels.

```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 4.5
yearly_general <- final_data %>%
  group_by(year) %>%
  summarize(avg_finish_hours = mean(finish_time_seconds)/3600, avg_temp = mean(avg_race_temp_f), group = "General Runner Sample", .groups="drop")
yearly_top1 <- top_1percent %>%
  group_by(year) %>%
  summarize(avg_finish_hours = mean(finish_time_seconds)/3600, avg_temp = mean(avg_race_temp_f), group = "Top 1% Runner Sample", .groups="drop")
plot_data1 <- bind_rows(yearly_general, yearly_top1)

ggplot(plot_data1, aes(x = avg_temp, y = avg_finish_hours, color = group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("General Runner Sample" = "steelblue", "Top 1% Runner Sample" = "orange")) +
  labs(title = "Thermal Sensitivity: General Sample vs. Top 1% Elites",
       x = "Average Race Temperature (°F)", y = "Average Finish Time (Hours)", color = "Runner Group") +
  theme_minimal() + theme(legend.position = "bottom")
```

## Weather Density Distribution
This density plot shows how the distribution of finish times shift under different weather conditions, categorized as cold, mild, and hot.
```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 4.5
final_data %>%
  mutate(temp_category = case_when(avg_race_temp_f < 50 ~ "Cold", 
                                   avg_race_temp_f < 65 ~ "Mild", 
                                   TRUE ~ "Hot")) %>%
  ggplot(aes(x = finish_time_seconds/3600, fill = temp_category)) +
  geom_density(alpha = 0.5) +
  labs(title = "Finish Time Density by Conditions", x = "Hours", y = "Density", fill = "Conditions") + theme_minimal()

```

## Quartile Sensitivity
This visualization shows how weather impacts runners differently depending on which performance quartile they are in.
```{r}
#| echo: false
#| fig-align: "center"
#| fig-height: 4.5
quartiles <- final_data %>%
  group_by(year, quartile) %>%
  summarize(avg_time_hours = mean(finish_time_seconds)/3600, avg_temp = mean(avg_race_temp_f), .groups = "drop")
ggplot(data=quartiles, aes(x=avg_temp, y=avg_time_hours, color=as.factor(quartile))) +
  geom_point() + geom_smooth(method="lm", se=FALSE) + scale_color_manual(values = c("1"="navy", "2"="blue", "3"="skyblue4", "4"="skyblue1"),
  labels = c("Q1 (Fastest)", "Q2", "Q3", "Q4 (Slowest)")) +
  labs(title = "Sensitivity by Performance Quartile", x = "Avg Temp (°F)", y = "Average Finish Time (Hours)", color = "Quartile") + theme_minimal()

```

# Key Findings

## 1. Heat Vs. Other Weather Elements
- **Finding:** Heat impacts runners in the Chicago Marathon more than any other weather element.
- **Proof:** Regression models confirm that apparent temperature and temperature explain a greater proportion of finish time variance ($R^2 \approx 0.50$ to $0.53$) compared to relative humidity. There was only 1 race with notable precipitation, and it was only 0.21 inches, so I would need more data to come to any conclusions.
```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 4
ggplot(plot_data1 %>% filter(group == "General Runner Sample"), aes(x = avg_temp, y = avg_finish_hours)) +
  geom_point(size = 2.5, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 1) +
  labs(title = "General Sample Finish Times vs. Temperature",
       x = "Average Temperature (°F)", y = "Finish Time (Hours)") +
  theme_minimal()
```

## 2. Finish Time per Degree
- **Finding:** For every 1°F increase in temperature, an average of `r round(gen_slope * 3600, 1)` seconds (`r round(gen_slope, 4)` hours) are added to a general runner's finish time.
- **Proof:** The linear regression slope coefficient from the general sample model shows what impact temperature has on the general marathon runner. 
```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 4
ggplot(plot_data1, aes(x = avg_temp, y = avg_finish_hours, color = group)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(values = c("General Runner Sample" = "steelblue", "Top 1% Runner Sample" = "orange")) +
  labs(title = "Thermal Sensitivity Slopes",
       x = "Average Temperature (°F)", y = "Finish Time (Hours)", color = "Group") +
  theme_minimal() + theme(legend.position = "bottom")
```

## 3. Running Groups Impacted by Heat
- **Major Insight:** Runners in the middle of the pack in quartiles 2 and 3 are impacted the most by heat.
- **Proof:** Quartiles 2 and 3 show higher individual $R^2$ values (Q2: 0.234, Q3: 0.264) than elites or the slowest quartile, which shows that their marathon performance is impacted the most by thermal stress.
```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 4
ggplot(data = quartiles, aes(x = avg_temp, y = avg_time_hours, color = as.factor(quartile))) +
  geom_point(alpha = 0.6, size = 2) +  
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +  
  scale_color_manual(values = c("1" = "navy", "2" = "blue", "3" = "skyblue4", "4" = "skyblue1"),
                     labels = c("Q1 (Fastest)", "Q2", "Q3", "Q4 (Slowest)")) +
  labs(title = "Temperature Sensitivity by Quartile",
       x = "Average Temperature (°F)", y = "Finish Time (Hours)", color = "Quartile") +  
  theme_minimal()
```


## 4. Ideal Racing Conditions
- **Major Insight:** The ideal racing conditions for the Chicago Marathon are in the 50-65°F temperature bracket.
- **Proof:** Density distributions across temperature ranges show that the marathon finish times are clustered tighter than other temperature ranges and they skew significantly faster compared the spikes with hotter weather.
```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 4
final_data %>%
  mutate(temp_category = case_when(
    avg_race_temp_f < 50 ~ "Cold (<50°F)",
    avg_race_temp_f < 65 & avg_race_temp_f >= 50 ~ "Mild (50-65°F)",
    TRUE ~ "Warm/Hot (>65°F)"
  )) %>%
  ggplot(aes(x = finish_time_seconds / 3600, fill = temp_category)) +
  geom_density(alpha = 0.5) +
  labs(title = "Finish Time Density Across Temperature Brackets",
       x = "Finish Time (Hours)", y = "Density", fill = "Conditions") +
  theme_minimal()
```

## Major Insights
- Temperature: Heat impacts runners more than any other weather element. It has a stronger relationship with finish time variance than precipitation or humidity.
- For every 1°F increase in temperature, an average of `r round(gen_slope, 2)` seconds are added to a general runner's finish time.

## Important Trends
- **General Population Slope:** `r round(gen_slope, 4)` sec/°F (95% CI: `r round(gen_ci$conf.low, 4)`–`r round(gen_ci$conf.high, 4)` sec/°F)
- **Elite Sample Slope:** `r round(elite_slope, 4)` sec/°F (95% CI: `r round(elite_ci$conf.low, 4)`–`r round(elite_ci$conf.high, 4)` sec/°F)
- **Quartile Patterns:** Runners from Q2 and Q3 had the highest R\^2 values (Q2: 0.234, Q3: 0.264). The weather conditions had a greater impact on the finishing times of runners from Q2 and Q3 than Q1 and Q4.
- **Performance Difference:** Elite runners experience a heat penalty ≈3 times less than the general population.

## Supporting Graphics

```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 5
yearly_general <- final_data %>%
  group_by(year) %>%
  summarize(avg_finish_hours = mean(finish_time_seconds)/3600, avg_temp = mean(avg_race_temp_f), group = "General Runner Sample", .groups="drop")

yearly_top1 <- top_1percent %>%
  group_by(year) %>%
  summarize(avg_finish_hours = mean(finish_time_seconds)/3600, avg_temp = mean(avg_race_temp_f), group = "Top 1% Runner Sample", .groups="drop")

plot_data1 <- bind_rows(yearly_general, yearly_top1)

ggplot(plot_data1, aes(x = avg_temp, y = avg_finish_hours, color = group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("General Runner Sample" = "steelblue", "Top 1% Runner Sample" = "orange")) +
  labs(title = "Thermal Sensitivity: General Sample vs. Top 1% Elites",
       subtitle = paste0("General Adj. R² = ", round(general_r2, 4), " | Elite Adj. R² = ", round(elite_r2, 4)),
       x = "Average Race Temperature (°F)", y = "Average Finish Time (Hours)", color = "Runner Group") +
  theme_minimal() + theme(legend.position = "bottom")
```

## Weather density distribution, finish times across temperature ranges

- Race conditions can be grouped as cold, mild, hot.
- Density curves show a skew.

```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 5
final_data %>%
  mutate(temp_category = case_when(
    avg_race_temp_f < 50 ~ "Cold (<50°F)",
    avg_race_temp_f < 65 & avg_race_temp_f >= 50 ~ "Mild (50-65°F)",
    TRUE ~ "Warm/Hot (>65°F)"
  )) %>%
  ggplot(aes(x = finish_time_seconds/3600, fill = temp_category)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Distribution Across Temperature Ranges",
       x = "Finish Time (Hours)", y = "Density", fill = "Conditions") +
  theme_minimal()
```

## Quartile Sensitivity Graphic

```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 4.5
quartiles <- final_data %>%
  group_by(year, quartile) %>%
  summarize(avg_time_hours = mean(finish_time_seconds)/3600, avg_temp = mean(avg_race_temp_f), .groups = "drop")

ggplot(data=quartiles, aes(x=avg_temp, y=avg_time_hours, color=as.factor(quartile))) +
  geom_point(alpha=0.6, size=2) + 
  geom_smooth(method="lm", se = FALSE, linewidth=1.2) + 
  scale_color_manual(values = c("1"="navy", "2"="blue", "3"="skyblue4", "4"="skyblue1"),
                     labels = c("Q1 (Fastest)", "Q2", "Q3", "Q4 (Slowest)")) +
  labs(title = "Temperature and Results by Quartile",
       x = "Average Temperature (°F)", y = "Average Finish Time (Hours)", color = "Quartile") + 
  theme_minimal()
```

## Temperature Distribution Graphic

```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 4.5
final_data %>%
  mutate(temp_category = case_when(
    avg_race_temp_f < 50 ~ "Cold (<50°F)",
    avg_race_temp_f < 65 & avg_race_temp_f >= 50 ~ "Mild (50-65°F)",
    TRUE ~ "Warm/Hot (>65°F)"
  )) %>%
  ggplot(aes(x = finish_time_seconds/3600, fill = temp_category)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Distribution Across Temperature Ranges",
       x = "Finish Time (Hours)", y = "Density", fill = "Conditions") +
  theme_minimal()
```

# Conclusions

## Implications

<div style="font-size: 50%;">
::: incremental
- **Apparent Temperature Superiority:** The apparent temperature explains a slightly higher proportion of finish time variance than the true temperature ($R^2 = 0.5284$ vs. $R^2 = 0.5023$). This supports the claim that environmental factors such as humidity and wind speed have an impact on physiological strain.
- **Heat Vulnerability By Gender:** There is no significant difference in how heat impacts men and women ($p = 0.4876$). Although women have a baseline offset ($p = 0.0053$), the heat impacts men (average of 4.28 hours) and women (average of 4.70 hours) at the same rate per each 1°F increase in temperature.
- **Weather Anomalies:** In 2007, there was a heat wave ($78^\circ\text{F}$–$88^\circ\text{F}$) which added an average of 24.5 to 27.5 minutes to the marathon finish times ($t = 34.778, p < 2.2 \times 10^{-16}$) compared to standard linear variance.
:::
</div>

## Recommendations

::: incremental
- **Assistance During Race:** There should be more stations available where runners can hydrate and cool off their bodies.
- **Medical Help:** Ahead of the race using API forecasts, the race organizers should prepare accordingly for potential injuries and help that would be needed during the race in case of emergency.
:::

## Future Work

::: incremental
- **Training:** A sample of the preparation and training each runner did ahead of the race would be valuable.\
- **Splits:** Taking splits every mile or so would be beneficial to analyze how different groups respond to general fatigue. These splits could also factor in the weather conditions during different parts of the race.
- Would it be beneficial to move the marathon to a different day of the year to make it more likely that the weather conditions are closer to the ideal race temperature?
:::
