---
title: "Chicago Marathon Analysis and the Impact of Weather on Performance"
author: "Miles Maloney"
institute: "STAT385"
date: "August 2, 2026"
format: 
  revealjs: 
    theme: default
editor: visual
#note: when hitting Render it does update on http://localhost:7093/#/title-slide
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

library(shiny)
library(tidyverse)
library(plotly)
library(DT)
library(broom)
library(leaflet)
library(sf)
#data
gpx_map_2000 <- st_read("2000 Chicago Marathon Route.gpx", layer = "tracks", quiet = TRUE)
gpx_map_2025 <- st_read("2025 Chicago Marathon Route.gpx", layer = "tracks", quiet = TRUE)
runner_data <- read_csv("STAT385 Project (2000 to 2025 results).csv")
weather_data <- read_csv("STAT385 Project (weather working set).csv")

#data wrangling
set.seed(1)
runners_finished <- runner_data %>%
  drop_na(finish_time_seconds) %>%
  filter(finish_time_seconds <= 23400) %>%
  group_by(year) %>%
  mutate(quartile=ntile(finish_time_seconds, 4)) %>%
  group_by(year, quartile) %>%
  slice_sample(n = 1000) %>%
  ungroup()

weather_needed <- weather_data %>%
  mutate(year = year(time), hour = hour(time))

racing_weather <- weather_needed %>%
  filter(hour >= 7 & hour <= 15) %>%
  group_by(year) %>%
  summarize(
    avg_race_temp_f = mean(temperature_2m_f, na.rm = TRUE), 
    avg_humidity_pct = mean(relative_humidity_2m_pct, na.rm = TRUE),
    total_precip_in = sum(precipitation_inch, na.rm = TRUE),
    .groups = "drop"
  )
final_data <- runners_finished %>%
  left_join(racing_weather, by = "year")
top_1percent <- runner_data %>%
  drop_na(finish_time_seconds) %>%
  filter(finish_time_seconds <= 23400) %>%
  group_by(year) %>%
  mutate(percentile = ntile(finish_time_seconds, 100)) %>%
  filter(percentile == 1) %>% 
  ungroup() %>%
  left_join(racing_weather, by ="year")

spread_data <- bind_rows(
  final_data %>% mutate(runner_group = "General Sample"),
  top_1percent %>% mutate(runner_group = "Top 1% Elites")
)

general_model <- lm(finish_time_seconds ~ avg_race_temp_f, data = final_data)
elite_model <- lm(finish_time_seconds ~ avg_race_temp_f, data = top_1percent)

general_r2 <- summary(general_model)$adj.r.squared
elite_r2 <- summary(elite_model)$adj.r.squared

gen_ci <- tidy(general_model, conf.int = TRUE, level = 0.95) %>% filter(term == "avg_race_temp_f")
elite_ci <- tidy(elite_model, conf.int = TRUE, level = 0.95) %>% filter(term == "avg_race_temp_f")

gen_slope <- gen_ci$estimate
elite_slope <- elite_ci$estimate  

```

# Motivation

## Why This Topic Matters

::: incremental
- The Chicago Marathon is one of the largest races in the world.
- Over 50,000 runners show up to participate in it in recent years.
- Runners spend months or years training for the Marathon.
- Weather on the race day is unpredictable.
- Understanding how runners are sensitive to the temperature conditions helps them set personal goals and helps marathon directors prepare for medical needs during the race.
:::

## Research Questions

::: incremental
- Does temperature impact the finish times of runners?
- Is there a difference in performance due to weather conditions between the General Sample and Elite Runners?
- Are casual runners or professional runners more sensative to weather changes?
- Do humidity or precipitation impact perfomance more than temperature?
:::

# Data Description

## Data Source

- The Marathon results were collected from a reliable public source as a CSV file. It contained results from all runners from 2000-2025, and had over 900,000 finishers.
- The hourly weather data was collected from Open-Meteo APIs, which provides high-resolution data from national weather services.
- Every year, the weather data was selected from the race window between 7am and 3pm. 7am was the earliest possible starting time, and 3pm was the latest possible finishing time.

## Variables

- The dependent variable is finish time. It was collected in seconds, but I converted it to hours.
- The independent variables are from the weather data, which are: Average Temperature (°F), Relative Humidity (%), and Total Precipitation (inches)
- The categorical variables are gender, age group, runner quartile, and runner group (general sample, top 1% of finishers)

## Data Size

- The raw data from race results was data from over 930,000 finishers across 25 years.\
- I took a stratified random sample of 1000 random runners in each quartile from each year. This was to make sure that each sample had the same size and to ensure efficiency.
- The top 1% of runners data from each year was collected and put in a different data set.

## Data Limitations

- Certain runner's heats started at different times.\
- There is no data on runner's training volume, diet, or potential personal challenges that could impact finish time.
- The model uses the average of the weather from 7am to 3pm.

# Methods

## Data Cleaning

- Filtered the data by the finishing time, so that all times were less than the Chicago Marathon time limit of 6 hours and 30 minutes (15-minute per mile pace).
- Removed runners who did not finish the race.
- Merged the hourly weather averages to the runner results based on `year`.

## Exploratory Analysis

::: incremental
- The change in marathon course details over the years was very important, so I used geographic data and sf to map out the course.
- I plotted the overall distribution of the marathon finishing times
- The distribution of marathon times is a broad bell shape. There are clusters before the finishing times of 3 hours, 4 hours, and 5 hours due to common goals that runners share.\
:::

## Marathon Finish Time Distribution

```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 5
#| fig-width: 10
ggplot(final_data, aes(x=finish_time_seconds/3600)) + 
  geom_histogram(aes(y=after_stat(density)), bins=100, color = "gray", fill="lightgray") + 
  geom_density(color ="blue", linewidth = 1) +
  labs(
    title = "Marathon Finish Times Distribution (2000-2025)",
    x = "Finishing Time (hours)",
    y = "Density"
  ) + 
  theme_minimal()
```

## Course Map Changes (2000 to 2025)

<p style="margin-top: -5px; font-size: 16px; color: #555;">

Interactive Chicago Marathon Course Map. Feel free to zoom in and select/unselect either year.

```{r}
#| echo: false
#| message: false
leaflet(height = "800px") %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolylines(data = gpx_map_2000, color = "orange", weight = 5, opacity = 0.9, 
               label = "2000 Course", group = "2000 Course") %>%
  addPolylines(data = gpx_map_2025, color = "blue", weight = 5, opacity = 0.9, 
               label = "2025 Course", group = "2025 Course") %>%
  addLayersControl(
    overlayGroups = c("2000 Course", "2025 Course"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

## Statistical Methods

- Used the linear regression model lm to find the relationship between weather factors and finishing times.
- Used the broom package to calculate the slopes and R\^2 values.

## Visualizations

- To visualize the trendlines, used ggplot
- Made interactive visuals with plotly, shiny, and leaflet so the viewer can interact with the data and change what they would like to look at and interact with.

# Key Findings

## Major Insights

## Important Trends

- **General Population Slope:** `r round(gen_slope, 4)` sec/°F (95% CI: `r round(gen_ci$conf.low, 4)`–`r round(gen_ci$conf.high, 4)` sec/°F)
- **Elite Sample Slope:** `r round(elite_slope, 4)` sec/°F (95% CI: `r round(elite_ci$conf.low, 4)`–`r round(elite_ci$conf.high, 4)` sec/°F)
- **Quartile Patterns:** Runners from Q2 and Q3 had the highest R\^2 values (Q2: 0.234, Q3: 0.264). The weather conditions had a greater impact on the finishing times of runners from Q2 and Q3 than Q1 and Q4.
- **Performance Difference:** Elite runners experience a heat penalty ≈3 times less than the general population.

## Supporting Graphics

```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 5
yearly_general <- final_data %>%
  group_by(year) %>%
  summarize(avg_finish_hours = mean(finish_time_seconds)/3600, avg_temp = mean(avg_race_temp_f), group = "General Runner Sample", .groups="drop")

yearly_top1 <- top_1percent %>%
  group_by(year) %>%
  summarize(avg_finish_hours = mean(finish_time_seconds)/3600, avg_temp = mean(avg_race_temp_f), group = "Top 1% Runner Sample", .groups="drop")

plot_data1 <- bind_rows(yearly_general, yearly_top1)

ggplot(plot_data1, aes(x = avg_temp, y = avg_finish_hours, color = group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("General Runner Sample" = "steelblue", "Top 1% Runner Sample" = "orange")) +
  labs(title = "Thermal Sensitivity: General Sample vs. Top 1% Elites",
       subtitle = paste0("General Adj. R² = ", round(general_r2, 4), " | Elite Adj. R² = ", round(elite_r2, 4)),
       x = "Average Race Temperature (°F)", y = "Average Finish Time (Hours)", color = "Runner Group") +
  theme_minimal() + theme(legend.position = "bottom")
```

## Weather density distribution, finish times across temperature ranges

- Race conditions can be grouped as cold, mild, hot.
- Density curves show a skew.

```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 5
final_data %>%
  mutate(temp_category = case_when(
    avg_race_temp_f < 50 ~ "Cold (<50°F)",
    avg_race_temp_f < 65 & avg_race_temp_f >= 50 ~ "Mild (50-65°F)",
    TRUE ~ "Warm/Hot (>65°F)"
  )) %>%
  ggplot(aes(x = finish_time_seconds/3600, fill = temp_category)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Distribution Across Temperature Ranges",
       x = "Finish Time (Hours)", y = "Density", fill = "Conditions") +
  theme_minimal()
```

## Quartile Sensitivity Graphic

```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 4.5
quartiles <- final_data %>%
  group_by(year, quartile) %>%
  summarize(avg_time_hours = mean(finish_time_seconds)/3600, avg_temp = mean(avg_race_temp_f), .groups = "drop")

ggplot(data=quartiles, aes(x=avg_temp, y=avg_time_hours, color=as.factor(quartile))) +
  geom_point(alpha=0.6, size=2) + 
  geom_smooth(method="lm", se = FALSE, linewidth=1.2) + 
  scale_color_manual(values = c("1"="navy", "2"="blue", "3"="skyblue4", "4"="skyblue1"),
                     labels = c("Q1 (Fastest)", "Q2", "Q3", "Q4 (Slowest)")) +
  labs(title = "Temperature and Results by Quartile",
       x = "Average Temperature (°F)", y = "Average Finish Time (Hours)", color = "Quartile") + 
  theme_minimal()
```

## Temperature Distribution Graphic

```{r}
#| echo: false
#| message: false
#| fig-align: "center"
#| fig-height: 4.5
final_data %>%
  mutate(temp_category = case_when(
    avg_race_temp_f < 50 ~ "Cold (<50°F)",
    avg_race_temp_f < 65 & avg_race_temp_f >= 50 ~ "Mild (50-65°F)",
    TRUE ~ "Warm/Hot (>65°F)"
  )) %>%
  ggplot(aes(x = finish_time_seconds/3600, fill = temp_category)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Distribution Across Temperature Ranges",
       x = "Finish Time (Hours)", y = "Density", fill = "Conditions") +
  theme_minimal()
```

# Conclusions

## Implications

## Recommendations

## \## Future Work
